---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2021"
author: "Chao Ouyang"
output:
  html_document:
    df_print: paged
subtitle: Assignment 3
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the project open the first thing you will do is change "Student Name" on line 3 with your name. 
Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A01_Sp21.Rmd"). Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the January 2021 Monthly Energy Review. Once again you will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. 

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r, warning=FALSE, message=FALSE}
# Load/install required package here
library(readxl)
library(ggplot2)
library(dplyr)
library(forecast)
library(tseries)
library(Kendall)
```

```{r}
# Import data
energy_data <- as.data.frame(read_excel(path = "../Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",
                                        skip = 10, sheet = 1, col_names = TRUE))
energy_data <- energy_data[-1, ]
rownames(energy_data) <- NULL

energy_data_sub <- energy_data[, c(1, 4, 5, 6)]
# str(energy_data_sub)
energy_data_sub[, 2:4] <- sapply(energy_data_sub[, 2:4], as.numeric)
# str(energy_data_sub)
# head(energy_data_sub, 10)

energy_data_sub$Month[1]
energy_data_sub$Month[nrow(energy_data_sub)]

ts_energy <- ts(energy_data_sub[, 2:4], start = c(1973, 1), end = c(2020, 10), frequency = 12)
head(ts_energy, 10)
```

## Trend Component

### Q1
Create a plot window that has one row and three columns. And then for each object on your data frame, fill the plot window with time series plot, ACF and PACF. You may use the some code form A2, but I want all three plots on the same window this time. (Hint: watch videos for M4)
```{r}
par(mfrow = c(1,3))

# 1
plot(ts_energy[, 1], type = "l", ylab = "Energy (Trillion BTU)", main = "Total Biomass \nEnergy Production")
abline(h = mean(ts_energy[, 1]), col = "red")

Acf(ts_energy[, 1], lag.max = 40, type = "correlation", plot = TRUE, main = "Total Biomass \nEnergy Production")
Pacf(ts_energy[, 1], lag.max = 40, plot = TRUE, main = "Total Biomass \nEnergy Production")

# 2
plot(ts_energy[, 2], type = "l", ylab = "Energy (Trillion BTU)", main = "Total Renewable \nEnergy Production")
abline(h = mean(ts_energy[, 2]), col = "red")

Acf(ts_energy[, 2], lag.max = 40, type = "correlation", plot = TRUE, main = "Total Renewable \nEnergy Production")
Pacf(ts_energy[, 2], lag.max = 40, plot = TRUE, main =  "Total Renewable \nEnergy Production")

# 3
plot(ts_energy[, 3], type = "l", ylab = "Energy (Trillion BTU)", main = "Hydroelectric Power \nConsumption")
abline(h = mean(ts_energy[, 3]), col = "red")

Acf(ts_energy[, 3], lag.max = 40, type = "correlation", plot = TRUE, main = "Hydroelectric Power \nConsumption")
Pacf(ts_energy[, 3], lag.max = 40, plot = TRUE, main = "Hydroelectric Power \nConsumption")
```



### Q2
From the plot in Q1, do the series Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?

##### The total biomass energy production series and the total renewable energy production series have more identifiable upward trends, which appear to be linear. It is also worth noting that their growth flattened over the period around the last decade of the 20th century. On the other hand, the hydroelectric power series seems to have a slight downward trend, whose linearity is not visually clear.



### Q3
Use the *lm()* function to fit a linear trend to the three time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.
```{r}
t <- c(1:nrow(ts_energy))

# 1
biomass_model <- lm(energy_data_sub$`Total Biomass Energy Production` ~ t)
summary(biomass_model)

# 2
re_model <- lm(energy_data_sub$`Total Renewable Energy Production` ~ t)
summary(re_model)

# 3
hydro_model <- lm(energy_data_sub$`Hydroelectric Power Consumption` ~ t)
summary(hydro_model)
```

##### The linear regression model on the biomass series yields an estimated intercept of 135.5 and slope of 0.47, with both being statistically significant. This means that the predicted biomass production starts at 135.5 trillion BTU in January 1973 and continues to rise at a rate of 0.47 trillion BTU per month until October 2020. The model explains 79.62% of the variation in the data.

##### The linear regression model on the renewable energy series yields an estimated intercept of 330.4 and slope of 0.84, with both being statistically significant. This means that the predicted renewable energy production starts at 330.4 trillion BTU in January 1973 and continues to rise at a rate of 0.84 trillion BTU per month until October 2020. The model explains 68.82% of the variation in the data.

##### The linear regression model on the hydropower series yields an estimated intercept of 258.1 and slope of -0.073, with both being statistically significant. This means that the predicted hydropower production starts at 258.1 trillion BTU in January 1973 and continues to decline at a rate of 0.073 trillion BTU per month until October 2020. The model explains 7.53% of the variation in the data.

 
 
### Q4
Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?
```{r}
detrended_data <- energy_data_sub
biomass_beta0 <- as.numeric(biomass_model$coefficients[1])
biomass_beta1 <- as.numeric(biomass_model$coefficients[2])
re_beta0 <- as.numeric(re_model$coefficients[1])
re_beta1 <- as.numeric(re_model$coefficients[2])
hydro_beta0 <- as.numeric(hydro_model$coefficients[1])
hydro_beta1 <- as.numeric(hydro_model$coefficients[2])

detrended_data$`Total Biomass Energy Production` <- energy_data_sub$`Total Biomass Energy Production` - 
  (biomass_beta0 + biomass_beta1 * t)
detrended_data$`Total Renewable Energy Production` <- energy_data_sub$`Total Renewable Energy Production` -
  (re_beta0 + re_beta1 * t)
detrended_data$`Hydroelectric Power Consumption` <- energy_data_sub$`Hydroelectric Power Consumption` -
  (hydro_beta0 + hydro_beta1 * t)

# 1
ggplot(energy_data_sub, aes(x = Month, y = energy_data_sub[, 2])) +
  geom_line(color = "blue") +
  ylab("Energy (Trillion BTU)") + xlab("Time") +
  ggtitle("Total Biomass Energy Production") + 
  geom_smooth(color = "red", method = "lm") +
  geom_line(aes(y = detrended_data[, 2]), col = "green") +
  geom_smooth(aes(y = detrended_data[, 2]), color = "gold", method = "lm") +
  theme_bw()

# 2
ggplot(energy_data_sub, aes(x = Month, y = energy_data_sub[, 3])) +
  geom_line(color = "blue") +
  ylab("Energy (Trillion BTU)") + xlab("Time") + 
  ggtitle("Total Renewable Energy Production") + 
  geom_smooth(color = "red", method = "lm") +
  geom_line(aes(y = detrended_data[, 3]), col = "green") +
  geom_smooth(aes(y = detrended_data[, 3]), color = "gold", method = "lm") +
  theme_bw()

# 3
ggplot(energy_data_sub, aes(x = Month, y = energy_data_sub[, 4])) +
  geom_line(color = "blue") +
  ylab("Energy (Trillion BTU)") + xlab("Time") + 
  ggtitle("Hydroelectric Power Consumption") + 
  geom_smooth(color = "red", method = "lm") +
  geom_line(aes(y = detrended_data[, 4]), col = "green") +
  geom_smooth(aes(y = detrended_data[, 4]), color = "gold", method = "lm") +
  theme_bw()
```

##### The upward trends in biomass and renewable energy production, and the downward trend in hydropower consumption are removed in the new plots. We see the new trendlines lying horizontal over time, even though the fluctuation patterns persist. 



### Q5
Plot ACF and PACF for the detrended series and compare with the plots from Q1. Did the plots change? How?
```{r}
par(mfrow = c(1,2))

detrended_ts_energy <- ts(detrended_data[, 2:4], start = c(1973, 1), end = c(2020, 10), frequency = 12)

# 1
Acf(detrended_ts_energy[, 1], lag.max = 40, type = "correlation", plot = TRUE, main = "Detrended Total Biomass \nEnergy Production")
Pacf(detrended_ts_energy[, 1], lag.max = 40, plot = TRUE, main = "Detrended Total Biomass \nEnergy Production")

# 2
Acf(detrended_ts_energy[, 2], lag.max = 40, type = "correlation", plot = TRUE, main = "Detrended Total Renewable \nEnergy Production")
Pacf(detrended_ts_energy[, 2], lag.max = 40, plot = TRUE, main =  "Detrended Total Renewable \nEnergy Production")

# 3
Acf(detrended_ts_energy[, 3], lag.max = 40, type = "correlation", plot = TRUE, main = "Detrended Hydroelectric Power \nConsumption")
Pacf(detrended_ts_energy[, 3], lag.max = 40, plot = TRUE, main = "Detrended Hydroelectric Power \nConsumption")
```

##### The ACF plots for biomass and renewable energy show a more visually identifiable annual pattern. The rest of the plots do not show obvious changes after comparison, except some slight changes in magnitudes.



## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6
Do the series seem to have a seasonal trend? Which serie/series? Use function *lm()* to fit a seasonal means model to this/these time series. Ask R to print the summary of the regression. Interpret the regression output. Save the regression coefficients for further analysis.

##### By visual observation, hydropower series seems to have a more noticable seasonal trend. 
```{r}
dummies <- seasonaldummy(ts_energy[, 1])

# fit seasonal means model
sm_biomass_model <- lm(energy_data_sub$`Total Biomass Energy Production` ~ dummies)
sm_re_model <- lm(energy_data_sub$`Total Renewable Energy Production` ~ dummies)
sm_hydro_model <- lm(energy_data_sub$`Hydroelectric Power Consumption` ~ dummies)

summary(sm_biomass_model)
summary(sm_re_model)
summary(sm_hydro_model)
```

##### Based on the regression results, the significance of many of the estimated season dummies in the seasonal means model for the hydropower series shows a seasonal trend for the series, which the other two series do not have. In the hydropower model, the intercept represents the estimated value for Dec, while the estimated coefficients for the dummies represent the difference between any of the dummy months and Dec. For example, we see that Mar-Jun have higher values than Dec, while Aug-Nov have lower values.



### Q7
Use the regression coefficients from Q6 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?
```{r}
beta_hydro_int <- sm_hydro_model$coefficients[1]
beta_hydro_coeff <- sm_hydro_model$coefficients[2:12]

# compute seasonal component
nobs <- nrow(energy_data_sub)
comp <- array(0, nobs)

for (i in 1:nobs){
  comp[i] <- beta_hydro_int + beta_hydro_coeff %*% dummies[i,]
}

colnames(energy_data_sub) <- gsub(" ", "_", colnames(energy_data_sub))
ggplot(energy_data_sub, aes(x = Month, y = Hydroelectric_Power_Consumption)) +
  geom_line(color = "blue") +
  geom_line(aes(y = comp), color = "red") +
  ylab("Energy (Trillion BTU)") + xlab("Time") + 
  ggtitle("Hydroelectric Power Consumption (with Seasonal Components)") +
  theme_bw()
  
deseasoned_hydro <- energy_data_sub$Hydroelectric_Power_Consumption - comp

ggplot(energy_data_sub, aes(x = Month, y = Hydroelectric_Power_Consumption)) +
  geom_line(color = "blue") +
  geom_line(aes(y = deseasoned_hydro), color = "green") +
  ylab("Energy (Trillion BTU)") + xlab("Time") + 
  ggtitle("Hydroelectric Power Consumption (Original vs. Deseasoned)") +
  theme_bw()
```

##### After seasonality is removed from the original hydropower series, we are able to observe more variations between months. The new series still shows a similar trend to the original series.


### Q8
Plot ACF and PACF for the deseason series and compare with the plots from Q1. Did the plots change? How?
```{r}
par(mfrow = c(1,2))

Acf(ts(deseasoned_hydro), lag.max = 40, type = "correlation", plot = TRUE, main = "Deseasoned Hydroelectric Power \nConsumption")
Pacf(ts(deseasoned_hydro), lag.max = 40, plot = TRUE, main = "Deseasoned Hydroelectric Power \nConsumption")
```

##### The ACF plot for the hydropower series changes drastically after deseasoning. We no longer observe negative ACF values and the seasonal pattern from the original ACF plot is much weaker now and decreasing gradually as lag increases. In the new PACF plot, we observe an overall decrease in magnitude compared to the origincal PACF plot.
